---
title: Analysis of the NYPD Shooting Incident Data
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

NYPD Shooting Incident Data contains every shooting incident that occurred in NYC since 2006 to 2020. This publicly available data is downloaded from <https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD>.

Each rqw in the dataset represents a shooting event, which contain information like Time of the incident, Location, details about the victim and the Suspect and also the victim demographics is included.

## Read Data

Data is read using the absolute path of the local folder. 

```{r read}
library(tidyverse)
nypd_dt <- read_csv("./NYPD_Shooting_Incident_Data__Historic_.csv")
```

## Tidy and Transform the Data

Goal of the Tidying step is to ensure that...
  
  1. Each variable is in a column
  2. Each observation is a row
  3. Each value is a cell

Furthermore, with the transform step we would look at the data types and add new columns and transformations as as necessary.

From the columns of interest perspective the data looks tidy. However, the data type of date column is 'chr', which should be converted to date type.

Following step will; 

1. Convert the data type of OCCUR_DATE to date. 
2. Remove the columns that is not required for the analysis
3. Add a new column derived from OCCUR_DATE called OCCUR_YEAR for easy analysis

There are missing data in some of the columns such as Age Group, Sex and race. These missing data with NA would be considered as a Unknown category in the Analysis.

```{r transform, echo=TRUE}
library(lubridate)
nypd_dt <- nypd_dt %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
  select(-c(Lon_Lat, Longitude, Latitude, X_COORD_CD, Y_COORD_CD))

nypd_dt <- nypd_dt %>%
  mutate(OCCUR_YEAR = year(OCCUR_DATE))

summary(nypd_dt)
```

## Analysis and visualization

### Analysis of Shooting Incidents by Year, Boro

```{r Analyze, echo=TRUE}
inc_by_year_boro <- nypd_dt %>%
  group_by(OCCUR_YEAR, BORO) %>%
  summarise(incidents = n()) %>%
  select (OCCUR_YEAR, BORO, incidents) %>%
  ungroup()

inc_by_year_boro
summary(inc_by_year_boro)
```

From the chart below it is observed that:

1. Bronx has the most shooting incidents
2. Shooting incidents declined since 2012 to 2019, but since 2019 there is a sharp increase.
3. Staten Island has the lowest incidents

```{r chart1, echo=TRUE}
ggplot(inc_by_year_boro, aes(OCCUR_YEAR, incidents, colour = BORO)) +
  geom_line() 
```


### Analysis of Shooting Incidents resulting in Death


```{r chart2, echo=TRUE}
NY_Incidents <- nypd_dt %>%
  group_by(month=floor_date(OCCUR_DATE, "month")) %>%
  summarise(incidents = n(), deaths = sum(STATISTICAL_MURDER_FLAG == TRUE)) %>%
  mutate(deaths_per_incident = deaths/incidents) %>%
  select(month, incidents, deaths, deaths_per_incident) %>%
  ungroup()

NY_Incidents
```

Following chart below shows the Shooting incidents across New York city and the relative
deaths by month.

```{r chart3, echo=TRUE}
NY_Incidents %>%
  ggplot(aes(x = month, y = incidents )) +
  geom_line(aes(color = "incidents")) +
  geom_point(aes(color= "incidents")) +
  geom_line(aes(y = deaths, color="deaths")) +
  geom_point(aes(y = deaths, color="deaths"))
```

### Modeling
A linear model is used to figure the relationship between shooting incidents and deaths
```{r model, echo=TRUE}
mod <- lm(deaths ~ incidents, data = NY_Incidents)
summary(mod)
```

### Use the Model to predict

 
```{r predict, echo=TRUE}
NY_Incidents_pred <- NY_Incidents %>% 
  mutate(pred = predict(mod))
NY_Incidents_pred
```

### Visualize the prediction

From the plot below, it seems like the relationship between shooting incidents and deaths are linear.
```{r vis, echo=TRUE}
NY_Incidents_pred %>% ggplot() +
  geom_point(aes(x = incidents, y = deaths ), color = "blue") +
  geom_point(aes(x = incidents, y = pred ), color = "red")
```


## Biases

A possible bias could be on the model selection. 






